---
title: "Lecture 09: Absolute Model fit and Path Analysis"
subtitle: ""
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2024-10-09"
date-modified: "2024-10-11"
sidebar: false
execute: 
  echo: true
  warning: false
output-location: default
code-annotations: below
highlight-style: "nord"
format: 
  uark-revealjs:
    scrollable: true
    chalkboard: true
    embed-resources: false
    code-fold: false
    number-sections: false
    footer: "ESRM 64503 - Lecture 09: Absolute Model fit and Path Analysis"
    slide-number: c/t
    tbl-colwidths: auto
    output-file: slides-index.html
  html: 
    page-layout: full
    toc: true
    toc-depth: 2
    toc-expand: true
    lightbox: true
    code-fold: false
    fig-align: center
filters:
  - quarto
  - line-highlight
---

```{=html}
<div class="card shadow">
    <div class="ml-3 mt-2">
        <svg xmlns="http://www.w3.org/2000/svg" width="54" height="14" viewBox="0 0 54 14">
            <g fill="none" fill-rule="evenodd" transform="translate(1 1)">
                <circle cx="6" cy="6" r="6" fill="#FF5F56" stroke="#E0443E" stroke-width=".5"></circle>
                <circle cx="26" cy="6" r="6" fill="#FFBD2E" stroke="#DEA123" stroke-width=".5"></circle>
                <circle cx="46" cy="6" r="6" fill="#27C93F" stroke="#1AAB29" stroke-width=".5"></circle>
            </g>
        </svg>
    </div>
    <div class="card-body">
        <h5 class="card-title">Today's Class</h5>
        <ul>
          <li>Multivariate linear models: an introduction</li>
          <li>How to form multivariate models in lavaan</li>
          <li>What parameters mean</li>
          <li>How they relate to the multivariate normal distribution</li>
        </ul>
    </div>
</div>
```
## Today's Class

## Homework 2: Setup

```{r}
#| output-location: default
library(ESRM64503)
library(kableExtra)
library(tidyverse)
library(DescTools) # Desc() allows you to quick screen data
library(lavaan) # Desc() allows you to quick screen data
# options(digits = 3)
head(dataMath)
dim(dataMath)
```

## Homework 2: Model 1

```{r}
#| output-location: column
model04.syntax = "
# Variances:
perf ~~ perf 
use ~~ use   
mas ~~ mas

# Covariance:
perf ~~ use
perf ~~ mas
use ~~ mas

# Means:
perf ~ 1  
use ~ 1   
mas ~ 1   
"

## Estimation for model01
model04.fit <- cfa(model04.syntax, data=dataMath, mimic="MPLUS", fixed.x = TRUE, estimator = "MLR") 

## Print output
standardizedsolution(model04.fit)
```

-   For Performance and Usefulness, there is a weak but significantly positive correlation with **r = .158 (.059), p = .007**

-   For Performance and Anxiety, there is a strong and significantly positive correlation with **r = .526 (.043), p \< .001**

-   For Usefulness and Anxiety, there is a moderate and significantly positive correlation with **r = .299 (.049), p \< .001**

## Homework 2: Model 2

```{r}
model04b.syntax = "
# Variances:
perf ~~ perf 
use ~~ use   
mas ~~ mas

# Covariance:
perf ~~ use
perf ~~ mas
use ~~ 0*mas

# Means:
perf ~ 1  
use ~ 1   
mas ~ 1   
"
model04b.fit <- cfa(model04b.syntax, data=dataMath, mimic="MPLUS", fixed.x = TRUE, estimator = "MLR") 

## Print output
standardizedsolution(model04b.fit) |> show_table()
```

-   For Performance and Usefulness, there is a weak and non-significantly positive correlation with **r = .017 (.056), p = .765**

-   For Performance and Anxiety, there is a strong and significantly positive correlation with **r = .523 (.044), p \< .001**

-   For Usefulness and Anxiety, there is a zero correlation with **r = 0 (0), p = NA**

## Homework 2: AIC and BIC

```{r}
anova(model04.fit, model04b.fit)
```

-   For Model 1, AIC = 6421.3 and BIC = 6456.0

-   For Model 2, AIC = 6446.5 and BIC = 6477.4

-   According to the criteria of lower AIC/BIC values indicating better model fit and the result of Likelihood ratio test ($\chi^2(0, 1)=29.031, p < .001$), we prefer model 1 as it has significantly better model fit than model 2.

# Path Analysis

-   Previous models considered correlations among observed variables (`perf` and `use`). These models may be limited in answering more complex research questions: whether A is a mediator between B and C

-   Path analysis: Multivariate Linear Models where Outcomes can be also Predictors

-   Path analysis details:

    -   Model identification

    -   Modeling workflow

-   Example Analyses

## Wrapping Up

1.  This lecture was an introduction to the estimation of multivariate linear models for multivariate outcomes the using path analysis/SEM package `lavaan`

2.  We saw that the model for continuous data uses the multivariate normal distribution in its likelihood function

### Next Class

1.  Does each model fit the data well (absolute model fit)?
2.  If not, how can we improve model fit?
3.  Which model fits better (relative model fit)?
4.  Answers are given in the following lectures...
