---
title: "Lecture 04: Distribution and Estimation"
author: "Jihong Zhang*, Ph.D; Jinbo He, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2024-09-16"
sidebar: false
execute: 
  echo: true
  warning: false
output-location: column
format: 
  html: 
    page-layout: full
    toc: true
    toc-depth: 2
    toc-expand: true
    lightbox: true
    code-fold: false
  uark-revealjs:
    scrollable: true
    chalkboard: true
    embed-resources: false
    code-fold: false
    number-sections: false
    footer: "ESRM 64503: Lecture 04: Distribution and Estimation"
    slide-number: c/t
    tbl-colwidths: auto
    output-file: slides-index.html
---

## Today's Class

-   [The building blocks]{style="color: tomato"}: **The basics of mathematical statistics:**

    -   Random variables: Definition & Types

    -   Univariate distribution

        -   General terminology (e.g., sufficient statistics)

        -   Univariate normal (aka Gaussian)

        -   Other widely used univariate distributions

    -   Types of distributions: Marginal \| Conditional \| Joint

    -   Expected values: means, variances, and the algebra of expectations

    -   Linear combinations of random variables

-   [The finished product]{style="color: violet"}: **How the GLM fits within statistics**

    -   The GLM with the normal distribution

    -   The statistical assumptions of the GLM

    -   How to assess these assumptions

# Unit 1: Random Variables & Statistical Distribution

## Definition of Random Variables

[**Random:**]{.underline} situations in which the certainty of the outcome is unkown and is at least in part due to chanvce

[**Variable:**]{.underline} a value that may change give the scope of a given problem or set of operations

[**Random Variable**]{.underline}: a variable whose outcome depends on chance (possible values might represent the possible outcomes of a yet-to-be performed experiment)

Today, we will denote a random variable with a lower-cased: ***x***

## Types of Random Variables

1.  **Continuous**
    -   Examples of continuous random variables:
        -   ***x*** represent the height of a person, draw at random
        -   *Y* (the outcome/DV in a GLM*)*
        -   Some variables like **exam score or motivation scores** are not "true" continuous variables, but it is convenient to consider them as "continuous"
2.  **Discrete** (also called categorical, generally)
    -   Example of discrete random variables:
        -   ***x*** represents the gender of a person, drawn at random
        -   *Y* (outcomes like yes/no; pass/not pass; master / not master a skill; survive / die)
3.  **Mixture of Continuous and Discrete**:
    -   Example of mixture: \begin{equation}
          x =
        \begin{cases}
          RT & \text{between 0 and 45 seconds} \\
          0 & \text{otherwise}
        \end{cases}       
        \end{equation}

## Key Features of Random Variable

1.  Random variables each are described by a **probability density / mass function (PDF) –** $f(x)$

    -   PDF indicates relative frequency of occurrence

    -   A PDF is a math function that gives rough picture of the distribution from which a random variable is draw

2.  The type of random variable dictates the name and nature of these functions:

    -   [Continuous random variables]{style="color: tomato"}:

        -   $f(x)$ is called a probability density function

        -   Area under curve must equal to 1 (found by calculus integration)

        -   Height of curve (the function value $f(x)$):

            -   Can be any positive number

            -   Reflects relative likelihood of an observation occurring

    -   Discrete random variables:

        -   $f(x)$ is called a probability mass function

        -   Sum across all values must equal 1

------------------------------------------------------------------------

::: {.callout-note style="font-size: 1.4em;"}
Both max and min values of temperature look like continuous random variables.
:::

```{r}
#| code-fold: true
#| output-location: default
#| column: screen
#| out-width: 100%
#| fig-align: center
#| fig-asp: .5
library(tidyverse)
temp <- read.csv("data/temp_fayetteville.csv")
temp$value_F <- (temp$value / 10 * 1.8) + 32
temp |> 
  ggplot() +
  geom_density(aes(x = value_F, fill = datatype), col = "white", alpha = .8) +
  labs(x = "Max/Min Temperature (F) at Fayetteville, AR (Sep-2023)", 
       caption = "Source: National Oceanic and Atmospheric Administration (https://www.noaa.gov/)") +
  scale_x_continuous(breaks = seq(min(temp$value_F), max(temp$value_F), by = 2)) +
  scale_fill_manual(values = c("tomato", "turquoise")) +
  theme_classic(base_size = 15) 
```

## Other key Terms

-   The sample space is the set of all values that a random variable x can take:
    -   **Example 1:** The sample space for a random variable x from a normal distribution $x \sim N(\mu_x, \sigma^2_x)$ is $(-\infty, +\infty)$.
    -   **Example 2:** The sample space for a random variable x representing the outcome of a coin flip is {H, T}
    -   **Example 3:** The sample space for a random variable x representing the outcome of a roll of a die is {1, 2, 3, 4, 5, 6}
-   When using generalized models, the trick is to pick a distribution with a sample space that matches the range of values obtainable by data

## Uses of Distributions in Data Analysis

-   Statistical models make distributional assumptions on various parameters and / or parts of data

-   These assumptions govern:

    -   How models are estimated

    -   How inferences are made

    -   How missing data may be imputed

-   If data do not follow an assumed distribution, inferences may be inaccurate

    -   Sometimes a very big problem, other times not so much

-   Therefore, it can be helpful to check distributional assumptions prior to running statistical analysis

## Continuous Univariate distributions

-   To demonstrate how continuous distributions work and look, we will discuss three:

    -   Uniform distribution

    -   Normal distribution

    -   Chi-square distribution

-   Each are described a set of parameters, which we will later see are what give us our inferences when we analysis data

-   What we then do is put constraints on those parameters based on hypothesized effects in data

## Uniform distribution

-   The uniform distribution is how to help set up how continuous distributions work

-   For a continuous random variable x that ranges from (a, b), the uniform probability density function is:

    $f(x) = \frac{1}{b-a}$

-   The uniform distribution has two parameters

    ```{r}
    x = seq(0, 3, .1)
    y = dunif(x, min = 0, max = 3)
    ggplot() +
      geom_point(aes(x = x, y = y)) +
      geom_path(aes(x = x, y = y)) +
      theme_bw()
    ```

## More on the Uniform Distribution

-   To demonstrate how PDFs work, we will try a few values:

```{r}
conditions <- tribble(
   ~x, ~a, ~b,
   .5,  0,  1,
  .75,  0,  1,
   15,  0, 20,
   15, 10, 20
) |> 
  mutate(y = dunif(x, min = a, max = b))
conditions
```

-   The uniform PDF has the feature that all values of ***x*** are equally likely across the sample space of the distribution
    -   Therefore, you do not see ***x*** in the PDF $f(x)$
-   The mean of the uniform distribution is $\frac{1}{2}(a+b)$
-   The variance of the uniform distribution is $\frac{1}{12}(b-a)^2$

## Univariate Normal Distribution

-   For a continuous random variable ***x*** (ranging from $-\infty$ to $\infty$), the univariate normal distribution function is:

$$
f(x) = \frac1{\sqrt{2\pi\sigma^2_x}}\exp(-\frac{(x-\mu_x)^2}{2\sigma^2_x})
$$

-   The shape of the distribution is governed by two parameters:

    -   The mean $\mu_x$

    -   The variance $\sigma^2_x$

    -   These parameters are called **sufficient statistics** (they contain all the information about the distribution)

-   The skewness (lean) and kurtosis (peakedness) are fixed

-   Standard notation for normal distributions is $x\sim N(\mu_x, \sigma^2_x)$

    -   Read as: "*x* follows a normal distribution with a mean $\mu_x$ and a variance $\sigma^2_x$"

-   Linear combinations of random variables following normal distributions result in a random variable that is normally distributed

## Univariate Normal Distribution in R: pnorm

Density (`dnorm`), distribution function (`pnorm`), quantile function (`qnorm`) and random generation (`rnorm`) for the normal distribution with mean equal to `mean` and standard deviation equal to `sd`.

```{r}
Z = seq(-5, 5, .1) # Z-score
ggplot() +
  aes(x = Z, y = pnorm(q = Z, lower.tail = TRUE)) +
  geom_point() +
  geom_path() +
  labs(x = "Z-score", y = "Cumulative probability",
       title = "`pnorm()` gives the cumulative probability function P(X < T)")
```

## Univariate Normal Distribution in R: dnorm

```{r}
x <- seq(-5, 5, .1)

params <- list(
  y_set1 = c(mu =  0, sigma2 = 0.2),
  y_set2 = c(mu =  0, sigma2 = 1.0),
  y_set3 = c(mu =  0, sigma2 = 5.0),
  y_set4 = c(mu = -2, sigma2 = 0.5)
)

y <- sapply(params, function(param) dnorm(x, mean = param['mu'], sd = param['sigma2']))

dt <- cbind(x, y) |> 
  as.data.frame() |> 
  pivot_longer(starts_with("y_"))

ggplot(dt) +
  geom_path(aes(x = x, y = value, color = name, group = name), linewidth = 1.3) +
  scale_color_manual(values = 1:4,
                     name = "",
                     labels = c('y_set1' = expression(mu*"=0, "*sigma^2*"=.02"),
                                'y_set2' = expression(mu*"=0, "*sigma^2*"=1.0"),
                                'y_set3' = expression(mu*"=0, "*sigma^2*"=5.0"),
                                'y_set4' = expression(mu*"=-2, "*sigma^2*"=0.5"))
                     ) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 13))
```

## Chi-Square Distribution

-   Another frequently used univariate distribution is the Chi-square distribution

    -   Sampling distribution of the variance follows a chi-square distribution

    -   Likelihood ratios follow a chi-square distribution

-   For a continuous random variable *x* (ranging from 0 to $\infty$), the chi-square distribution is given by:

    $$
    f(x) =\frac1{2^{\frac{\upsilon}{2}} \Gamma(\frac{\upsilon}{2})} x^{\frac{\upsilon}{2}-1} \exp(-\frac{x}2)
    $$

-   $\Gamma(\cdot)$ is called the gamma function

-   The chi-square distribution is govern by one parameter: $\upsilon$ (the degrees of freedom)

    -   The mean is equal to $\upsilon$; the variance is equal to 2$\upsilon$

```{r}
x <- seq(0.01, 15, .01)
df <- c(1, 2, 3, 5, 10)
dt2 <- as.data.frame(sapply(df, \(df) dchisq(x, df = df)))
dt2_with_x <- cbind(x = x, dt2)
dt2_with_x |> 
  pivot_longer(starts_with("V")) |> 
  ggplot() +
  geom_path(aes(x = x, y = value, color = name), linewidth = 1.2) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_discrete(name = "", labels = paste("df =", df)) +
  labs(y = "f(x)") +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 13))
```
