---
title: "Lecture 04: Distribution and Estimation"
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2024-09-16"
sidebar: false
execute: 
  echo: true
  warning: false
output-location: column
format: 
  html: 
    page-layout: full
    toc: true
    toc-depth: 2
    toc-expand: true
    lightbox: true
    code-fold: false
  uark-revealjs:
    scrollable: true
    chalkboard: true
    embed-resources: false
    code-fold: false
    number-sections: false
    footer: "ESRM 64503: Lecture 04: Distribution and Estimation"
    slide-number: c/t
    tbl-colwidths: auto
    output-file: slides-index.html
filters:
  - shinylive
---

## Today's Class

-   Review Homework 1

-   [The building blocks]{style="color: tomato"}: **The basics of mathematical statistics:**

    -   Random variables: Definition & Types

    -   Univariate distribution

        -   General terminology (e.g., sufficient statistics)

        -   Univariate normal (aka Gaussian)

        -   Other widely used univariate distributions

    -   Types of distributions: Marginal \| Conditional \| Joint

    -   Expected values: means, variances, and the algebra of expectations

    -   Linear combinations of random variables

-   [The finished product]{style="color: violet"}: **How the GLM fits within statistics**

    -   The GLM with the normal distribution

    -   The statistical assumptions of the GLM

    -   How to assess these assumptions
    
# ESRM 64503: Homework 1

## Question 2

<details>

- Copy and paste your R syntax and R output that calculates the group **Senior-Old's standard error of group mean** (use the data and model we've used in class). 

  - **Aim**: Test whether the group mean of senior-old significantly higher than the baseline (here, 0 score)

$$
\mathbf{Test = \beta_0 +\beta_1Senior + \beta_2New +\beta_3Senior*New}
$$

- The group mean of Senior-Old is $\beta_0 + \beta_1$.

- Based on algebra for variance, we know hat

$$
Var(\beta_0 + \beta_1) = Var(\beta_0)+Var(\beta_1) +2*Cov(\beta_0, \beta_1)
$$

- `vcor` function provides the variances and covaraicnes for $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$

```{r}
#| output-location: default
# R syntax
library(ESRM64503)
library(tidyverse)
model1 <- lm(Test ~ Senior + New + Senior * New, data = dataTestExperiment)
data("dataTestExperiment")
Var_beta_0 <- vcov(model1)[1,1]
Var_beta_1 <- vcov(model1)[2,2]
Cov_beta_0_1 <- vcov(model1)[1,2]
Var_Test_SeniorOld <- Var_beta_0 + Var_beta_1 + 2*Cov_beta_0_1
(SE_Test_SeniorOld <- sqrt(Var_Test_SeniorOld)) 
```

</details>

## Question 3

<details>

- Copy and paste your R syntax and R output that calculates the standard error of conditional main effect of New (New vs. Old) when Senior = 1.

  - **Aim**: Test whether the conditional main effect of New when Senior significantly different from 0
     - In other words, when individuals are senior, whether new or old instruction method has significantly differences in their test scores


$$
\beta_{1|New =1} = \beta_1 + \beta_3 * 1
$$


- Based on algebra for variance, we know hat


$$
Var(\beta_1 + \beta_3) = Var(\beta_1)+Var(\beta_3) +2*Cov(\beta_1, \beta_3)
$$


-   `vcor` function provides the variances and covaraicnes for $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ 


```{r}
#| output-location: default
# R syntax
Var_beta_1 <- vcov(model1)[2,2]
Var_beta_3 <- vcov(model1)[4,4]
Cov_beta_1_3 <- vcov(model1)[2,4]
Var_betaofSenior_New1 <- Var_beta_1 + Var_beta_3 + 2*Cov_beta_1_3
(SE_betaofSenior_New1 <- sqrt(Var_betaofSenior_New1)) 
```
</details>

# Unit 1: Random Variables & Statistical Distribution

## Definition of Random Variables

- [**Random:**]{.underline} situations in which the certainty of the outcome is unkown and is at least in part due to chance

- [**Variable:**]{.underline} a value that may change give the scope of a given problem or set of operations

- [**Random Variable**]{.underline}: a variable whose outcome depends on chance (possible values might represent the possible outcomes of a yet-to-be performed experiment)

Today, we will denote a random variable with a lower-cased: ***x***

## Types of Random Variables

1.  **Continuous**
    -   Examples of continuous random variables:
        -   ***x*** represent the height of a person, draw at random
        -   *Y* (the outcome/DV in a GLM)
        -   Some variables like **exam score or motivation scores** are not "true" continuous variables, but it is convenient to consider them as "continuous"
2.  **Discrete** (also called categorical, generally)
    -   Example of discrete random variables:
        -   ***x*** represents the gender of a person, drawn at random
        -   *Y* (outcomes like yes/no; pass/not pass; master / not master a skill; survive / die)
3.  **Mixture of Continuous and Discrete**:
    -   Example of mixture: \begin{equation}
          x =
        \begin{cases}
          RT & \text{between 0 and 45 seconds} \\
          0 & \text{otherwise}
        \end{cases}       
        \end{equation}

## Key Features of Random Variable

1.  Random variables each are described by a **probability density / mass function (PDF) –** $f(x)$

    -   PDF indicates relative frequency of occurrence

    -   A PDF is a math function that gives rough picture of the distribution from which a random variable is draw

2.  The type of random variable dictates the name and nature of these functions:

    -   [Continuous random variables]{style="color: tomato"}:

        -   $f(x)$ is called a probability density function

        -   Area under curve must equal to 1 (found by calculus integration)

        -   Height of curve (the function value $f(x)$):

            -   Can be any positive number

            -   Reflects relative likelihood of an observation occurring

    -   [Discrete random variables]{style="color: violet"}:

        -   $f(x)$ is called a probability mass function

        -   Sum across all values must equal 1

------------------------------------------------------------------------

::: {.callout-note style="font-size: 1.4em;"}
Both max and min values of temperature can be considered as continuous random variables.
:::

- Question 1: what are the probabilities of integrating all values of max and min temperatures ({1, 1} or {0.5, 0.5})

- **Answer**: <span class="heimu"> it is {1, 1} for max and min temperatures. Because they are two seperated random variables.</span>


```{r}
#| code-fold: true
#| output-location: default
#| fig-align: center
#| fig-asp: .5
library(tidyverse)
temp <- read.csv("data/temp_fayetteville.csv")
temp$value_F <- (temp$value / 10 * 1.8) + 32
temp |> 
  ggplot() +
  geom_density(aes(x = value_F, fill = datatype), col = "white", alpha = .8) +
  labs(x = "Max/Min Temperature (F) at Fayetteville, AR (Sep-2023)", 
       caption = "Source: National Oceanic and Atmospheric Administration (https://www.noaa.gov/)") +
  scale_x_continuous(breaks = seq(min(temp$value_F), max(temp$value_F), by = 5)) +
  scale_fill_manual(values = c("tomato", "turquoise")) +
  theme_classic(base_size = 13) 
```

## Other key Terms

-   The sample space is the set of all values that a random variable x can take:
    -   **Example 1:** The sample space for a random variable x from a normal distribution $x \sim N(\mu_x, \sigma^2_x)$ is $(-\infty, +\infty)$.
    -   **Example 2:** The sample space for a random variable x representing the outcome of a coin flip is {H, T}
    -   **Example 3:** The sample space for a random variable x representing the outcome of a roll of a die is {1, 2, 3, 4, 5, 6}
-   When using generalized models, the trick is to pick a distribution with a sample space that matches the range of values obtainable by data
    - Logistic regression - Match Bernoulli distribution <span class="heimu">(Example 2)</span>
    - Poisson regression - Match Poisson distribution <span class="heimu">(Example 3)</span>

## Uses of Distributions in Data Analysis

-   Statistical models make distributional assumptions on various parameters and / or parts of data

-   These assumptions govern:

    -   How models are estimated

    -   How inferences are made

    -   How missing data may be imputed

-   If data do not follow an assumed distribution, inferences may be inaccurate

    -   Sometimes a very big problem, other times not so much

-   Therefore, it can be helpful to check distributional assumptions prior to running statistical analysis

## Continuous Univariate distributions

-   To demonstrate how continuous distributions work and look, we will discuss three:

    -   Uniform distribution

    -   Normal distribution

    -   Chi-square distribution

-   Each are described a set of parameters, which we will later see are what give us our inferences when we analysis data

-   What we then do is put constraints on those parameters based on hypothesized effects in data

## Uniform distribution

-   The uniform distribution is how to help set up how continuous distributions work
    
    - Typically, used for simulation studies that parameters are randomly generated

-   For a continuous random variable x that ranges from (a, b), the uniform probability density function is:

    $f(x) = \frac{1}{b-a}$

-   The uniform distribution has two parameters

    ```{r}
    x = seq(0, 3, .1)
    y = dunif(x, min = 0, max = 3)
    ggplot() +
      geom_point(aes(x = x, y = y)) +
      geom_path(aes(x = x, y = y)) +
      theme_bw()
    ```

------------------------------------------------------------------------

```{shinylive-r}
#| standalone: true
#| viewerHeight: 800
library(shiny)
library(bslib)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(1234)
# Define UI for application that draws a histogram
ui <- fluidPage(
  
  # Application title
  titlePanel("Uniform distribution"),
  
  # Sidebar with a slider input for number of bins 
  sidebarLayout(
    sidebarPanel(
      sliderInput("a",
                  "Lower Bound (a):",
                  min = 1,
                  max = 20,
                  value = 1,
                  animate = animationOptions(interval = 5000, loop = TRUE)),
      uiOutput("b_slider")
    ),
    
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("distPlot")
    )
  )
)

# Define server logic required to draw a histogram
x <- seq(0, 40, .02)
server <- function(input, output) {
  observeEvent(input$a, {
    output$b_slider <<- renderUI({
      sliderInput("b",
                  "Upper Bound (b):",
                  min = as.numeric(input$a),
                  max = 40,
                  value = as.numeric(input$a) + 1)
    })
    # browser()
    y <<- reactive({dunif(x, min = as.numeric(input$a), max = as.numeric(input$b))})
  })
  # browser()
  observe({
    output$distPlot <- renderPlot({
      # generate bins based on input$bins from ui.R
      ggplot() +
        aes(x = x, y = y())+
        geom_point() +
        geom_path() +
        labs(x = "x", y = "probability") +
        theme_bw() +
        theme(text = element_text(size = 20))
    })
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```


## More on the Uniform Distribution

-   To demonstrate how PDFs work, we will try a few values:

```{r}
conditions <- tribble(
   ~x, ~a, ~b,
   .5,  0,  1,
  .75,  0,  1,
   15,  0, 20,
   15, 10, 20
) |> 
  mutate(y = dunif(x, min = a, max = b))
conditions
```

-   The uniform PDF has the feature that all values of ***x*** are equally likely across the sample space of the distribution
    -   Therefore, you do not see ***x*** in the PDF $f(x)$
-   The mean of the uniform distribution is $\frac{1}{2}(a+b)$
-   The variance of the uniform distribution is $\frac{1}{12}(b-a)^2$

## Univariate Normal Distribution

-   For a continuous random variable ***x*** (ranging from $-\infty$ to $\infty$), the univariate normal distribution function is:

$$
f(x) = \frac1{\sqrt{2\pi\sigma^2_x}}\exp(-\frac{(x-\mu_x)^2}{2\sigma^2_x})
$$

-   The shape of the distribution is governed by two parameters:

    -   The mean $\mu_x$

    -   The variance $\sigma^2_x$

    -   These parameters are called **sufficient statistics** (they contain all the information about the distribution)

-   The skewness (lean) and kurtosis (peakedness) are fixed

-   Standard notation for normal distributions is $x\sim N(\mu_x, \sigma^2_x)$

    -   Read as: "*x* follows a normal distribution with a mean $\mu_x$ and a variance $\sigma^2_x$"

-   Linear combinations of random variables following normal distributions result in a random variable that is normally distributed

## Univariate Normal Distribution in R: pnorm

Density (`dnorm`), distribution function (`pnorm`), quantile function (`qnorm`) and random generation (`rnorm`) for the normal distribution with mean equal to `mean` and standard deviation equal to `sd`.

```{r}
Z = seq(-5, 5, .1) # Z-score
ggplot() +
  aes(x = Z, y = pnorm(q = Z, lower.tail = TRUE)) +
  geom_point() +
  geom_path() +
  labs(x = "Z-score", y = "Cumulative probability",
       title = "`pnorm()` gives the cumulative probability function P(X < T)")
```

## Univariate Normal Distribution in R: dnorm

```{r}
x <- seq(-5, 5, .1)

params <- list(
  y_set1 = c(mu =  0, sigma2 = 0.2),
  y_set2 = c(mu =  0, sigma2 = 1.0),
  y_set3 = c(mu =  0, sigma2 = 5.0),
  y_set4 = c(mu = -2, sigma2 = 0.5)
)

y <- sapply(params, function(param) dnorm(x, mean = param['mu'], sd = param['sigma2']))

dt <- cbind(x, y) |> 
  as.data.frame() |> 
  pivot_longer(starts_with("y_"))

ggplot(dt) +
  geom_path(aes(x = x, y = value, color = name, group = name), linewidth = 1.3) +
  scale_color_manual(values = 1:4,
                     name = "",
                     labels = c('y_set1' = expression(mu*"=0, "*sigma^2*"=.02"),
                                'y_set2' = expression(mu*"=0, "*sigma^2*"=1.0"),
                                'y_set3' = expression(mu*"=0, "*sigma^2*"=5.0"),
                                'y_set4' = expression(mu*"=-2, "*sigma^2*"=0.5"))
                     ) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 13))
```

## Chi-Square Distribution

-   Another frequently used univariate distribution is the Chi-square distribution

    -   Sampling distribution of the variance follows a chi-square distribution

    -   Likelihood ratios follow a chi-square distribution

-   For a continuous random variable *x* (ranging from 0 to $\infty$), the chi-square distribution is given by:

    $$
    f(x) =\frac1{2^{\frac{\upsilon}{2}} \Gamma(\frac{\upsilon}{2})} x^{\frac{\upsilon}{2}-1} \exp(-\frac{x}2)
    $$

-   $\Gamma(\cdot)$ is called the gamma function

-   The chi-square distribution is govern by one parameter: $\upsilon$ (the degrees of freedom)

    -   The mean is equal to $\upsilon$; the variance is equal to 2$\upsilon$

------------------------------------------------------------------------
```{r}
x <- seq(0.01, 15, .01)
df <- c(1, 2, 3, 5, 10)
dt2 <- as.data.frame(sapply(df, \(df) dchisq(x, df = df)))
dt2_with_x <- cbind(x = x, dt2)
dt2_with_x |> 
  pivot_longer(starts_with("V")) |> 
  ggplot() +
  geom_path(aes(x = x, y = value, color = name), linewidth = 1.2) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_discrete(name = "", labels = paste("df =", df)) +
  labs(y = "f(x)") +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 13))
```

------------------------------------------------------------------------

```{shinylive-r}
#| standalone: true
#| viewerHeight: 800
library(shiny)
library(bslib)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(1234)
# Define UI for application that draws a histogram
ui <- fluidPage(

    # Application title
    titlePanel("Chi-square distribution"),

    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
            sliderInput("Df",
                        "Degree of freedom:",
                        min = 1,
                        max = 20,
                        value = 1,
                        animate = animationOptions(interval = 5000, loop = TRUE)),
           verbatimTextOutput(outputId = "Df_display")
        ),

        # Show a plot of the generated distribution
        mainPanel(
           plotOutput("distPlot")
        )
    )
)

# Define server logic required to draw a histogram
server <- function(input, output) {
    x <- seq(0.01, 15, .01)
    df <- reactive({input$Df}) 
    
    output$Df_display <- renderText({
        paste0("DF = ", df())
    })
    output$distPlot <- renderPlot({
        # generate bins based on input$bins from ui.R
        
        
        dt2 <- as.data.frame(sapply(df(), \(df) dchisq(x, df = df)))
        dt2_with_x <- cbind(x = x, dt2)
        dt2_with_x |> 
            pivot_longer(starts_with("V")) |> 
            ggplot() +
            geom_path(aes(x = x, y = value), linewidth = 1.2) +
            scale_y_continuous(limits = c(0, 1)) +
            labs(y = "f(x)") +
            theme_bw() +
            theme(legend.position = "top", text = element_text(size = 13))
    })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

# Marginal, Joint, And Conditional Distribution

## Moving from One to Multiple Random Variables

-   When more than one random variable is present, there are several different types of statistical distributions:

-   We will first consider two discrete random variables：

    -   *x* is the outcome of the flip of a penny {$H_p$, $T_p$}

        -   $f(x=H_p) = .5$; $f(x =T_p) = .5$

    -   *z* is the outcome of the flip of a dime {$H_d$, $T_d$}

        -   $f(z=H_p) = .5$; $f(z =T_p) = .5$

-   We will consider the following distributions:

    -   Marginal distribution

        -   The distribution of one variable only (either $f(x)$ or $f(z)$)

    -   Joint distribution

        -   $f(x, z)$: the distribution of both variables (both *x* and *z*)

    -   **Conditional distribution**

        -   The distribution of one variable, conditional on values of the other:

            -   $f(x|z)$: the distribution of *x* given *z*

            -   $f(z|x)$: the distribution of *z* given x

## Marginal Distribution

-   Marginal distributions are what we have worked with exclusively up to this point: they represent the distribution by itself

    -   Continuous univariate distributions

    -   Categorical distributions

        -   The flip of a penny

        -   The flip of a dime

## Joint Distribution

-   Joint distributions describe the distribution of more than one variable, simultaneously

    -   Representations of multiple variables collected

-   Commonly, the joint distribution function is denoted with all random variables separated by commas

    -   In our example, $f(x,z)$ is the joint distribution of the outcome of flipping both a penny and a dime

        -   As both are discrete, the joint distribution has four possible values:

            \(1\) $f(x = H_p,z=H_d)$ (2) $f(x = H_p,z=T_d)$ (3) $f(x = T_p,z=H_d)$ (4) $f(x = T_p,z=T_d)$

-   Joint distributions are **multivariate distributions**

-   We will use joint distributions to introduce two topics

    -   Joint distributions of independent variables

    -   Joint distributions – used in maximum likelihood estimation

## Joint Distributions of Independent Random Variables

-   Random variables are said to be independent if the occurrence of one event makes it neither more nor less probable of another event

    -   For joint distributions, this means: $f(x,z)=f(x)f(z)$

-   In our example, flipping a penny and flipping a dime are independent – so we can complete the following table of their joint distribution:

    |                 | z = $H_d$                          | z = $T_d$                         | Marginal (Penny)              |
    |------------------|-------------------|------------------|------------------|
    | $x = H_p$       | $\color{tomato}{f(x=H_p, z=H_d)}$  | $\color{tomato}{f(x=H_p, z=T_d)}$ | $\color{turquoise}{f(z=H_p)}$ |
    | $x = T_p$       | $\color{tomato}{f(x= T_p, z=H_d)}$ | $\color{tomato}{f(x=T_p, z=T_d)}$ | $\color{turquoise}{f(z=T_d)}$ |
    | Marginal (Dime) | $\color{turquoise}{f(z=H_d)}$      | $\color{turquoise}{f(z=T_d)}$     |                               |

## Joint Distributions of Independent Random Variables

-   Because the coin flips are independent, this because:

|                 | z = $H_d$                            | z = $T_d$                           | Marginal (Penny)              |
|-----------------|-------------------|-------------------|-----------------|
| $x = H_p$       | $\color{tomato}{f(x=H_p)f( z=H_d)}$  | $\color{tomato}{f(x=H_p)f( z=T_d)}$ | $\color{turquoise}{f(z=H_p)}$ |
| $x = T_p$       | $\color{tomato}{f(x= T_p)f( z=H_d)}$ | $\color{tomato}{f(x=T_p)f( z=T_d)}$ | $\color{turquoise}{f(z=T_d)}$ |
| Marginal (Dime) | $\color{turquoise}{f(z=H_d)}$        | $\color{turquoise}{f(z=T_d)}$       |                               |

-   Then, with numbers:

|                 | z = $H_d$               | z = $T_d$               | Marginal (Penny)        |
|------------------|------------------|------------------|------------------|
| $x = H_p$       | $\color{tomato}{.25}$   | $\color{tomato}{.25}$   | $\color{turquoise}{.5}$ |
| $x = T_p$       | $\color{tomato}{.25}$   | $\color{tomato}{.25}$   | $\color{turquoise}{.5}$ |
| Marginal (Dime) | $\color{turquoise}{.5}$ | $\color{turquoise}{.5}$ |                         |

## Marginalizing Across a Joint Distribution

-   If you had a joint distribution, $\color{orchid}{f(x, z)}$, but wanted the marginal distribution of either variable ($f(x)$ or $f(z)$) you would have to **marginalize** across one dimension of the joint distribution.

-   For [**categorical random variables**]{.underline}, marginalize = sum across every value of z

$$
f(x) = \sum_zf(x, z)
$$

-   For example, $f(x = H_p) = f(x = H_p, z=H_d) +f(x = H_p, z=T_d)=.5$

-   For [continuous random variables,]{.underline} marginalize = integrate across z

    -   The integral:

        $$
        f(x) = \int_zf(x,z)dz
        $$

## Conditional Distributions

-   For two random variables x and z, a conditional distribution is written as: $f(z|x)$

    -   The distribution of z given x

-   The conditional distribution is equal to the joint distribution divided by the marginal distribution of the conditioning random variable

    $$
    f(z|x) = \frac{f(z,x)}{f(x)}
    $$

-   Conditional distributions are found everywhere in statistics

    -   The general linear model uses the conditional distribution variable

        $$
        Y \sim N(\beta_0+\beta_1X, \sigma^2_e)
        $$

## Example: Conditional Distribution

-   For two discrete random variables with {0, 1} values, the conditional distribution can be shown in a contingency table:

|                 | z = $H_d$               | z = $T_d$               | Marginal (Penny)        |
|------------------|------------------|------------------|------------------|
| $x = H_p$       | $\color{tomato}{.25}$   | $\color{tomato}{.25}$   | $\color{turquoise}{.5}$ |
| $x = T_p$       | $\color{tomato}{.25}$   | $\color{tomato}{.25}$   | $\color{turquoise}{.5}$ |
| Marginal (Dime) | $\color{turquoise}{.5}$ | $\color{turquoise}{.5}$ |                         |

Conditional: $f(z | x= H_p)$:

$f(z=H_d|x =H_p) = \frac{f(z=H_d, x=H_p}{f(x = H_p)} = \frac{.25}{.5}=.5$

$f(z = T_d | x = H_p)= \frac{f(z=T_d, x=H_p}{f(x=H_p)} = \frac{.25}{.5} = .5$

# Expected Values and The Algebra of Expectation

## Expected Values

-   Expected values are statistics taken the sample space of a random variable: they are essentially weighted averages

    ```{r}
    #| output-location: default
    set.seed(1234)
    x = rnorm(100, mean = 0, sd = 1)
    weights = dnorm(x, mean = 0, sd = 1)
    mean(weights * x)
    ```

-   The weights used in computing this average correspond to the probabilities (for a discrete random variable) or to the densities (for a continuous random variable)

::: {.callout-note font-size="1.2em"}
The expected value is represented by $E(x)$

The actual statistic that is being weighted by the PDF is put into the parenthesis where x is now
:::

-   Expected values allow us to understand what a statistical model implies about data, for instance:

    -   How a GLM specifies the (conditional) mean and variance of a DV

## Expected Value Calculation

-   For discrete random variables, the expected value is found by:

    $$
    E(x) = \sum_x xP(X=x)
    $$

-   For example, the expected value of a roll of a die is:

    $$
    E(x) = (1)\frac16+ (2)\frac16+(3)\frac16+(4)\frac16+(5)\frac16+6\frac16
    $$

-   For continuous random variables, the expected value is found by

    $$
    E(x) = \int_x xf(x)dx
    $$

-   We won't be calculating theoretical expected values with calculus... we use them only to see how models imply things about out data

## Variance and Covariance... As Expected Values

-   A distribution's theoretical variance can also be written as an expected value:

    $$
    V(x) = E(x-E(x))^2 = E(x -\mu_x)^2
    $$

    -   This formula helps us understand predictions made by GLMs and how that corresponds to statistical parameters we interpret

-   For a roll of a die, the theoretical variance is:

    $$
    V(x) = E(x - 3.5)^2 = \frac16(1-3.5)^2 + \frac16(2-3.5)^2 + \frac16(3-3.5)^2 + \frac16(4-3.5)^2 + \frac16(5-3.5)^2 + \frac16(6-3.5)^2 = 2.92
    $$

-   Likewise, for a pair of random variable *x* and *z*, the covariance can be found from their joint distributions:

    $$
    \text{Cov}(x,z)=E(xz)-E(x)E(z) = E(xz)-\mu_x\mu_z
    $$

    ```{r}
    #| results: hold
    set.seed(1234)
    N = 100
    x = rnorm(N, 2, 1)
    z = rnorm(N, 3, 1)
    xz = x*z
    (Cov_xz = mean(xz)-mean(x)*mean(z)) / ((N-1)/N) # unbiased covariance
    cov(x, z)
    ```

# Linear Combination of Random Variables

## Linear Combinations of Random Variables

-   A linear combination is an expression constructed from a set of terms by multiplying each term by a constant and then adding the results $$
    x = c + a_1z_1+a_2z_2+a_3z_3 +\cdots+a_nz_n
    $$

-   More generally, linear combinations of random variables have specific implications for the mean, variance, and possibly covariance of the new random variable

-   As such, there are predicable ways in which the means, variances, and covariances change

## Algebra of Expectations

-   Sums of Constants

$$
E(x+c) = E(x)+c \\
\text{Var}(x+c) = \text{Var}(x) \\
\text{Cov}(x+c, z) = \text{Cov}(x, z)
$$

### Example

Imagine for weight variable, each individual increases 3 lbs:

```{r}
#| output-location: default
set.seed(1234)
library(ESRM64503)
x = dataSexHeightWeight$weightLB
c_ = 3
z = dataSexHeightWeight$heightIN
mean(x + c_); mean(x)+c_
var(x + c_); var(x)
cov(x, z); cov(x+c_, z) # decimal place issue, near() accepts two values' diff less than .00001
```

------------------------------------------------------------------------

-   Products of Constants:

$$
E(cx) = cE(x) \\
\text{Var}(cx) = c^2\text{Var}(x) \\
\text{Cov}(cx, dz) = c*d*\text{Cov}(x, z)
$$
Imagine you wanted to convert weight from pounds to kilograms (where 1 pound = .453 kg) and convert height from inches to cm (where 1 inch = 2.54 cm)


```{r}
#| output-location: default

c_ = .453
mean(x*c_); mean(x)*c_
var(x*c_); var(x)*c_^2

d_ = 2.54
cov(c_*x, d_*z);c_*d_*cov(x, z)
```

------------------------------------------------------------------------

-   Sums of Multiple Random Variables:

$$
E(cx+dz) = cE(x) + dE(z) \\
\text{Var}(cx+dz) = c^2\text{Var}(x) + d^2\text{Var}(z) + 2c*d*\text{Cov}(x,z)\\
$$

```{r}
#| output-location: default
mean(x*c_+z*d_); mean(x)*c_+mean(z)*d_
var(x*c_+z*d_); c_^2*var(x) + d_^2*var(z) + 2*c_*d_*cov(x, z)
```

## Where We Use This Algebra

- Remember how we calculated the standard error of conditional main effect from simple main effect and interaction effect

$$
\mathbf{Test = 82.20 + 2.16*Senior + 7.76*New - 3.04*Senior*New}
$$

- Conditional Main Effects of Senior When New is 1 is (2.16 - 3.04) = -0.88

    - We know that $Var(\beta_{Senior})$ is 0.575,  $Var(\beta_{Senior*New})$ is 1.151, and $Cov(\beta_{Senior}, \beta_{Senior*New})$ is -0.575.

```{r}
#| output-location: default
model1 <- lm(Test ~ Senior + New + Senior * New, data = dataTestExperiment)
vcov(model1)
```

Then, 

$$
\mathbf{Var(\beta_{Senior}+\beta_{Senior*New}) = Var(\beta_{Senior}) + Var(\beta_{Senior*New}) + 2Cov(\beta_{Senior},\beta_{Senior*New})
\\ = 0.575 + 1.151 - 2* 0.575 = 0.576} 
$$
Thus, $SE(\beta_{Senior}+\beta_{Senior*New})=0.759$

## Combining the GLM with Expections

- Using the algebra of expectations predicting Y from X and Z:

$$
\hat{Y}_p=E(Y_p)=E(\beta_0+\beta_1X_p+\beta_2Z_p+\beta_3X_pZ_p+e_p) \\
= \beta_0+\beta_1X_p+\beta_2Z_p+\beta_3X_pZ_p+E(e_p) \\
= \beta_0+\beta_1X_p+\beta_2Z_p+\beta_3X_pZ_p
$$

- The variance of $f(Y_p|X_p, Z_p)$:

$$
V(Y_P)=V(\beta_0+\beta_1X_p+\beta_2Z_p+\beta_3X_pZ_p+e_p)\\ = V(e_p)=\sigma_e^2
$$

## Examining What This Means in the Context of Data

- If you recall from the regression analysis of the height/weight data, the final model we decided to interpret: Model 5

$$
W_p = \beta_0+\beta_1 (H_p-\bar{H}) +\beta_2F_p+\beta_3 (H_p-\bar{H}) F_p + e_p
$$

where $e_p \sim N(0, \sigma_e^2)$

```{r}
#| output-location: default
dat <- dataSexHeightWeight
dat$heightIN_MC <- dat$heightIN - mean(dat$heightIN)
dat$female <- dat$sex == 'F'
mod5 <- lm(weightLB ~ heightIN_MC + female + female*heightIN_MC, data = dat)
summary(mod5)
```

## Picturing the GLM with Distributions

- The distributional assumptions of the GLM are the reason why we do not need to worry if our dependent variable is normally distributed

- Our dependent variable should be conditionally normal

- We can check this assumption by checking our assumption about the residuals,
$e_p \sim N(0, \sigma^2_e)$ 

```{r}
#| echo: false
#| output-location: default
library(ggridges)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(1234)
pred_y <- predict(mod5, newdata = dat)
pred_y_dat <- data.frame(Point = names(pred_y), weightLB = pred_y, 
                         heightIN_MC = dat$heightIN_MC)
sigma <- sd(residuals(mod5))

residual_dat <- as.data.frame(sapply(pred_y, \(x) rnorm(100, mean = x, sd = sigma)))

residual_dat_long <- residual_dat |> 
  pivot_longer(everything())

colnames(residual_dat_long) <- c("Point", "weightLBVaribant")

residual_dat_long <- residual_dat_long |> 
  left_join(pred_y_dat) |> 
  arrange(Point) 

ggplot() +
  geom_density_ridges(aes(x = weightLBVaribant, y = as.factor(round(heightIN_MC, 3))), 
                      data = residual_dat_long, rel_min_height = 0.005, alpha = .5,
                      scale = .7) + 
  geom_point(aes(col = female, x = weightLB, y = as.factor(round(heightIN_MC, 3))), data= dat) +
  geom_smooth(aes(col = female, group = female, 
                  x = weightLB, y = as.factor(round(heightIN_MC, 3))), 
              method = 'lm', se = FALSE, data= dat) +
  labs(y = "HeightLB", x = "WeightLB") +
  theme_classic() +
  coord_flip()
```


## Assessing Distributional Assumptions Graphically

```{r}
plot(mod5)
```

## Hypothesis Tests for Normality

If a given test is **significant**, then it is saying that your data do not come from a normal distribution

In practice, test will give diverging information quite frequently:
the best way to evaluate normality is to consider both plots and tests (approximate = good)

```{r}
shapiro.test(mod5$residuals)
```

## Wrapping Up

- Today was an introduction to mathematical statistics as a way to understand the implications statistical models make about data

- Although many of these topics do not seem directly relevant, they help provide insights that untrained analysts may not easily attain