---
title: "DiagrammeR package"
author: Jihong Zhang
date: "2025-03-24"
fig-cap-location: top  
execute: 
  eval: true
format: html
keep-md: true
---

::: objectives
## Overview {.unnumbered}

- This section is based on the official [documentation](https://rich-iannone.github.io/DiagrammeR/) of `DiagrammeR`.
- This html widget is developed by Richard Iannone. 
- The DiagrammeR package allows create, modify, analyze, and visualize **network graph** diagrams
:::

## Installation

```{r}
#| eval: false
devtools::install_github("rich-iannone/DiagrammeR")
```

or 

```{r}
#| eval: false
install.packages("DiagrammeR")
```

## Start from scratch

Use `create_graph` to start with a empty graph object and use `add_node()` and `add_edge` functions to add nodes and edges, respectively.

```{r}
library(DiagrammeR)

create_graph() |> 
  add_node() |> 
  add_node() |> 
  add_edge(from = 1, to = 2) |> 
  render_graph(layout = "nicely")
```

## aesthetic for node and edge

Use `node_aes()` inside `add_node` and `edge_aes()` inside `add_edge` to specify aesthetic attributes.

```{r}
#| echo: false
library(tidyverse)

tribble(
  ~`Aesthetic`, ~Attributes,
  "Node", "color", 
  "", "fillcolor", 
  "", "fontcolor", 
  "Edge", "color", 
  "", "arrowhead", 
  "", "tooltip"
) |> 
  kableExtra::kable()
```
