---
execute: 
  warning: false
  message: false
format: html
---

### plotly package

-   This section also includes [Chapter 2 - Overview](https://plotly-r.com/overview) in [Book: Interactive web-based data visualization with R, plotly, and shiny](https://plotly-r.com/), which has more details about `plotly` package.

#### Basic Information

`plotly` R package is powered by the Javascript library *plotly.js*. A `plotly` object is initialized with a `plot_ly()` function.

```{r}
library(plotly)
library(ggplot2)
head(diamonds)
```

-   `plot_ly()` tries to find a sensible geometric representation of that information for

-   Note that column name is formatted with `~Name` in `plot_ly()`, which is different from `ggplot2` package `Name` or `vars(Name)`.

```{r}
#| layout-ncol: 1
#| fig-cap:
#|   - "Plot 1"
#|   - "Plot 2"
#|   - "Plot 3"
plot_ly(diamonds, x=~cut)
plot_ly(diamonds, x=~cut, y=~clarity, colors="Accent")
plot_ly(diamonds, x=~cut, color=~clarity, colors="Accent")
```

-   To specify the color directly, use `I()` function to declar the color value as "asis".

```{r}
plot_ly(diamonds, x=~cut, 
        color=I("tomato"), 
        stroke=I("royalblue"), 
        span = I(3)) |> 
layout(title = "Histogram with I() function")
```

#### Translate ggplot2 to plotly

In @fig-plotly-1, `ggplotly()` can translate a `ggplot2` object into a plotly object.

```{r}
#| label: fig-plotly-1
p <- ggplot(diamonds, aes(x=log(carat), y=log(price))) + 
  geom_hex(bins = 100)
ggplotly(p)
```

#### `add_` as layers and `|>` as connection

Plotly implements geoms in a similar fashion as ggplot2, functions that start in add\_ add a layer to the plot (e.g.: add_lines, add_bars), making it easy to combine series into a single chart, as in @fig-plotly-2.

```{r}
#| label: fig-plotly-2
#| fig-cap: "Plotly with add_ and |> as connection"
plot_ly(mtcars, x=~disp) |> 
  add_markers(y=~mpg, text = rownames(mtcars)) |> 
  add_lines(y=~fitted(loess(mpg ~ disp)))
```

#### Data in plotly.js

![A diagram of what happens when you print a plotly graph](https://plotly-r.com/images/printing.svg)

- Use `plotly_json()` to see the JSON representation of the plotly object.

- Use `jasonlite` to quickly check jason elements. Check out the [package information](https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html).

```{r}
library(jsonlite)
library(tidyverse)
p <- plot_ly(diamonds, x = ~cut, color = ~clarity, colors = "Accent")
p_jason <- plotly_json(p, FALSE, pretty = TRUE) |> 
  fromJSON() 
colnames(p_jason$data)
```
#### Work directly with plotly.js

`add_trace(type = "scatter", mode = "xxx")` function allows you to work directly with plotly.js. This function is a low-level interface to plotly.js, compared to other functions like `add_markers`.

```{r}
set.seed(99)
plot_ly() |> 
 add_trace(
   type = "scatter",
   mode = "markers+lines+text",
   x = 4:6, 
   y = 4:6,
   text = replicate(3, praise::praise("You are ${adjective}! 🙌")),
   textposition = "right",
   hoverinfo = "text",
   textfont = list(family = "MiSans", size = 16)
 ) |> 
 layout(xaxis = list(range = c(3, 8)))
```



