---
title: "Lecture 07: Generalized Linear Models (Binary Outcome) and Matrix Algebra"
subtitle: "Matrix Algebra in R"
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2024-09-24"
sidebar: false
execute: 
  echo: true
  warning: false
output-location: column
code-annotations: below
format: 
  uark-revealjs:
    scrollable: true
    chalkboard: true
    embed-resources: false
    code-fold: false
    number-sections: false
    footer: "ESRM 64503 - Lecture 07: Matrix Algebra"
    slide-number: c/t
    tbl-colwidths: auto
    output-file: slides-index.html
  html: 
    page-layout: full
    toc: true
    toc-depth: 2
    toc-expand: true
    lightbox: true
    code-fold: false
    fig-align: center
filters:
  - quarto
  - line-highlight
---

## Today's Class

-   Matrix Algebra
-   Multivariate Normal Distribution
-   Multivariate Linear Analysis

## Graduate Certificate

# An Introduction to Matrix Algebra

## Matrix Elements

-   A matrix (denote as capitalized **X**) is composed of a set of elements

    -   Each element is denote by its position in the matrix (row and column)

```{r}
X = matrix(c(
  1, 2,
  3, 4,
  5, 6
), nrow = 3, byrow = TRUE)
X
```

```{r}
dim(X) # Number of rows and columns
```

-   In R, we use `matrix[rowIndex, columnIndex]` to extract the element with the position of rowIndex and columnIndex

```{r}
X[2, 1]
```

-   In statistics, we use $x_{ij}$ to represent one element with the position of *i*th row and *j*th column. For a example matrix $\mathbf{X}$ with the size of 1000 rows and 2 columns.

    -   The first subscript is the index of the rows

    -   The second subscript is the index of the columns

$$
\mathbf{X} = \begin{bmatrix}
x_{11} & x_{12}\\
x_{21} & x_{22}\\
\dots &  \dots \\
x_{1000, 1} & x_{1000,2}
\end{bmatrix}
$$

## Scalars

-   A scalar is just a single number

-   The name scalar is important: the number "scales" a vector – it can make a vector "longer" or "shorter".

-   Scalars are typically written without boldface:

    $$
    x_{11} = 520
    $$

-   Each element of a matrix is a scalar.

-   Matrices can be multiplied by scalar so that each elements are multiplied by this scalar

    ```{r}
    3 * X
    ```

## Matrix Transpose

-   The transpose of a matrix is a reorganization of the matrix by switching the indices for the rows and columns

    $$
    \mathbf{X} = \begin{bmatrix}
    520 & 580\\
    520 & 550\\
    \vdots &  \vdots\\
    540 & 660\\
    \end{bmatrix}
    $$

$$
\mathbf{X}^T = \begin{bmatrix}
520 & 520 & \cdots & 540\\
580 & 550 & \cdots & 660
\end{bmatrix}
$$

-   An element $x_{ij}$ in the original matrix $\mathbf{X}$ is now $x_{ij}$ in the transposed matrix $\mathbf{X}^T$

-   **Transposes are used to align matrices for operations where the sizes of matrices matter (such as matrix multiplication)**

    ```{r}
    t(X)
    ```

## Types of Matrices

-   **Square Matrix:** A square matrix has the same number of rows and columns

    -   Correlation / covariance matrices are square matrices

-   **Diagonal Matrix**: A diagonal matrix is a square matrix with non-zero diagonal elements ($x_{ij}\neq0$ for $i=j$) and zeros on the off-diagonal elements ($x_{ij} =0$ for $i\neq j$):

    $$
    \mathbf{A} = \begin{bmatrix}
    2.758 & 0 & 0 \\
    0 & 1.643 & 0 \\
    0 & 0     & 0.879\\
    \end{bmatrix}
    $$

    -   We will use diagonal matrices to form correlation matrices
