---
title: "Exercise 04: Data Manipulation and Function Scope Exercises"
subtitle: ""
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2025-08-18"
sidebar: false
execute: 
  warning: false
  message: false
  eval: false
  echo: true
format: 
  html: 
    page-layout: full
toc: true
toc-depth: 2
lightbox: true
code-fold: show
---


# Data Import Exercises

## Exercise 0: Importing Data from Website

You are working on downloading the [UARK Final exam table](https://registrar.uark.edu/registration/final-exam-schedule/spring-2025-final-exam-schedule.php) and import it into R for data analysis.

- Use `rvest` package to construct a data table

```{r}
#| eval: false
#| echo: false
#| message: false 
html_text_raw <- '
<table class="table table-bordered">
                  
                  <tbody>
                     
                     <tr>
                        
                        <th>Final Exam Date</th>
                        
                        <th>Class Meeting Time/<br>Common Final</th>
                        
                        <th>Exam Time</th>
                        </tr>
                     
                     <tr>
                        
                        <td rowspan="7">Monday,<br>May 5, 2025</td>
                        
                        <td>MATH 11103, 12003, 21003, 22003</td>
                        
                        <td>During Posted Math Testing Center Hours</td>
                        </tr>
                     
                     <tr>
                        
                        <td>CHEM 12103; CHEM 14103</td>
                        
                        <td>8:00 AM to 10:00 AM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 10:45 AM</td>
                        
                        <td>10:15 AM to 12:15 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 12:55 PM</td>
                        
                        <td>12:45 PM to 2:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 2:00 PM</td>
                        
                        <td>3:00 PM&nbsp; to 5:00 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>SEVI 30103/SEVI 301H3 <br>MATH 13004, 24005, 24004, 240H4, 25004, 250H4, 26004, 260H4</td>
                        
                        <td>5:30 PM to 7:30 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 5:15 PM <br>Mon &amp; Mon/Wed Classes Starting<br>5:00 PM or Later Alternate Time Slot for Makeup Exams</td>
                        
                        <td>7:45 PM to 9:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td rowspan="8">Tuesday,<br>May 6, 2025</td>
                        </tr>
                     
                     <tr>
                        
                        <td>MATH 11003 (Sections 51 and Above) <br>See Instructor for Details</td>
                        
                        <td>During Posted Math Testing Center Hours</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 9:30 AM</td>
                        
                        <td>8:00 AM to 10:00 AM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>FOREIGN LANGUAGE 10103, 10203, 10106, 11003, 20103, 20203<br>CLASSES (FREN, JAPN, CHIN) EXCEPT ARAB 10103, 10106, 20106; <br>CHRK 10103, 10203; GERM 10103, 10203, 20103, 20203; <br>GREK 10203, 20103, 20203; <br>ITAL 10103, 10203, 20103, 20203;<br>LATN 10103, 10203, 20203; <br>RUSS 10103, 10203, 20103, 20203;<br>SPAN 10103, 10203, 20103, 20203, 202H3</td>
                        
                        <td>10:15 AM to 12:15 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 12:30 PM</td>
                        
                        <td>12:45 PM to 2:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 3:30 PM</td>
                        
                        <td>3:00 PM&nbsp; to 5:00 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>FINN 20403/FINN 204H3</td>
                        
                        <td>5:30 PM to 7:30 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue &amp; Tue/Thu <br>Classes Starting 5:00 PM or Later<br>Alternate Time Slot for Makeup Exams</td>
                        
                        <td>7:45 PM to 9:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td rowspan="7">Wednesday,<br>May 7, 2025</td>
                        
                        <td>MATH 11003 (Sections 1 - 50)<br>See Instructor for Details</td>
                        
                        <td>During Posted Math Testing Center Hours</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 8:35 AM</td>
                        
                        <td>8:00 AM to 10:00 AM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 9:40 AM</td>
                        
                        <td>10:15 AM to 12:15 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 11:50 AM</td>
                        
                        <td>12:45 PM to 2:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed/Fri 3:05 PM</td>
                        
                        <td>3:00 PM&nbsp; to 5:00 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>SEVI 20503/SEVI 205H3 <br>CHEM 14203/CHEM 142H3 <br>CHEM 36203/CHEM 362H3/CHEM 37103</td>
                        
                        <td>5:30 PM to 7:30 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Mon/Wed 4:35 <br>Mon/Wed/Fri 4:10 PM <br>Wed Classes Starting 5:00 PM or Later <br>Alternate Time Slot for Makeup Exams</td>
                        
                        <td>7:45 PM to 9:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td rowspan="7">Thursday,<br>May 8, 2025</td>
                        
                        <td>MATH 20503</td>
                        
                        <td>During Posted Math Testing Center Hours</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 8:00 AM</td>
                        
                        <td>8:00 AM to 10:00 AM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 11:00 AM</td>
                        
                        <td>10:15 AM to 12:15 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 2:00 PM</td>
                        
                        <td>12:45 PM to 2:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Tue/Thu 4:00 PM <br>Tue/Thu 4:30 PM</td>
                        
                        <td>3:00 PM to 5:00 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>PHYS 20203<br>MGMT 21003/MGMT 210H3</td>
                        
                        <td>5:30 PM to 7:30 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>Thu Classes Starting 5:00 PM or Later Alternate Time Slot for Makeup Exams</td>
                        
                        <td>7:45 PM to 9:45 PM</td>
                        </tr>
                     
                     <tr>
                        
                        <td rowspan="4">Friday,<br>May 9, 2025</td>
                        
                        <td>Mon/Wed/Fri 7:30 AM Alternate Time Slot for Makeup Exams</td>
                        
                        <td>8:00 AM to 10:00 AM</td>
                        </tr>
                     
                     <tr>
                        
                        <td>ISYS 21003/ISYS 210H3; <br>BUSI 10303/BUSI 103H3<br>Alternate Time Slot for Makeup Exams</td>
                        
                        <td>10:15 AM to 12:15 PM</td>
                        </tr>
                     </tbody>
                  </table>
'
final_example_tbl <- html_table(html_element(minimal_html(html_text_raw), "table"))
final_example_tbl
```


## Exercise 1: Importing Data from CSV

You are working with the following CSV file containing employee information:

```
Employee ID,Full Name,Department,Salary,Years of Experience
1,John Doe,HR,50000,5
2,Jane Smith,Engineering,N/A,8
3,Michael Johnson,Marketing,55000,6
4,Susan Lee,HR,52000,4
5,David Kim,Engineering,65000,7
6,Amy Brown,Marketing,57000,5
```

1. Use the `read_csv()` function to import this data into R.
  - For illustration, you can use the literal data (hint: use `I()`) to import the csv text.

```{r}
#| eval: false
#| echo: false
#| message: false 
library(readr)
employee_data <- read_csv(I(
"
Employee ID,Full Name,Department,Salary,Years of Experience
1,John Doe,HR,50000,5
2,Jane Smith,Engineering,N/A,8
3,Michael Johnson,Marketing,55000,6
4,Susan Lee,HR,52000,4
5,David Kim,Engineering,65000,7
6,Amy Brown,Marketing,57000,5
"  
))
employee_data
```  
  
2. Clean the data by replacing any missing salary values with NA.

```{r}
#| eval: false
#| echo: false
#| message: false 
employee_data_NATrans <- read_csv(I(
"
Employee ID,Full Name,Department,Salary,Years of Experience
1,John Doe,HR,50000,5
2,Jane Smith,Engineering,N/A,8
3,Michael Johnson,Marketing,55000,6
4,Susan Lee,HR,52000,4
5,David Kim,Engineering,65000,7
6,Amy Brown,Marketing,57000,5
"  
), na = "N/A")
employee_data_NATrans
```

3. Rename the columns to use snake case (employee_id, full_name, department, salary, years_of_experience).

```{r}
#| eval: false
#| echo: false
#| message: false 
library(dplyr)
employee_data_cleaned <- employee_data_NATrans %>%
  rename(employee_id = `Employee ID`,
         full_name = `Full Name`,
         department = `Department`,
         salary = `Salary`,
         years_of_experience = `Years of Experience`)
employee_data_cleaned
```





## Exercise 2: Importing Data Using Custom Delimiters

You have another dataset where the values are separated by semicolons (;) instead of commas:

```
Employee ID;Full Name;Department;Salary;Years of Experience
1;John Doe;HR;50000;5
2;Jane Smith;Engineering;60000;8
3;Michael Johnson;Marketing;55000;6
4;Susan Lee;HR;52000;4
5;David Kim;Engineering;65000;7
6;Amy Brown;Marketing;57000;5
```

- Use `read_delim()` to import this dataset with the semicolon delimiter (`;`).

```{r}
#| eval: false
#| echo: false
#| message: false
data_txt = "
Employee ID;Full Name;Department;Salary;Years of Experience
1;John Doe;HR;50000;5
2;Jane Smith;Engineering;60000;8
3;Michael Johnson;Marketing;55000;6
4;Susan Lee;HR;52000;4
5;David Kim;Engineering;65000;7
6;Amy Brown;Marketing;57000;5"
employee_data_semicolon <- read_delim(I(data_txt), delim = ";")
employee_data_semicolon
```

# Data Tidying Exercises

## Exercise 3: Pivoting Data Longer

You have a dataset that tracks the number of customers visiting different stores each week:

```{r}
store_visits <- tibble(
  store = c("Store A", "Store B", "Store C"),
  week_1 = c(120, 150, 110),
  week_2 = c(130, 160, 120),
  week_3 = c(140, 170, 130),
  week_4 = c(110, 140, 100),
  week_5 = c(160, 180, 150)
)
```

- Use `pivot_longer()` to convert this dataset into a long format, where each row represents a single observation for a specific week.

```{r}
#| eval: false
#| echo: false
#| message: false
store_visits_long <- store_visits %>%
  pivot_longer(
    cols = starts_with("week_"), 
    names_to = "week", 
    values_to = "visits"
  )
store_visits_long
```

## Exercise 4: Pivoting Data Wider

You have a long-format dataset that tracks the average test scores of students in different subjects over several terms:

```{r}
scores_long <- tibble(
  student = c("Alice", "Alice", "Bob", "Bob", "Charlie", "Charlie"),
  term = c("Term 1", "Term 2", "Term 1", "Term 2", "Term 1", "Term 2"),
  subject = c("Math", "Math", "Math", "Math", "Math", "Math"),
  score = c(85, 90, 78, 80, 92, 95)
)
```


- Use `pivot_wider()` to convert this dataset into a wide format, where each term (e.g., "Term 1", "Term 2") is a column, and the scores are the values.

```{r}
#| eval: false
#| echo: false
#| message: false
scores_wide <- scores_long %>%
  pivot_wider(
    names_from = term, 
    values_from = score
  )
scores_wide
```


## Exercise 5: Combining pivot_longer() and pivot_wider()

You have a dataset tracking the number of calls received by a customer service center, split by department and month:

```{r}
calls_wide <- tibble(
  department = c("Sales", "Support", "Billing"),
  january = c(200, 150, 180),
  february = c(210, 160, 190),
  march = c(220, 170, 200)
)
```


- First, use `pivot_longer()` to convert this dataset into a long format, where each row corresponds to a specific department and month.
- Then, use `pivot_wider()` to convert the long dataset back into a wide format, where the months are columns and the values represent the number of calls.

```{r}
#| eval: false
#| echo: false
#| message: false
calls_long <- calls_wide %>%
  pivot_longer(
    cols = c("january", "february", "march"), 
    names_to = "month", 
    values_to = "calls"
  )
calls_wide_again <- calls_long %>%
  pivot_wider(
    names_from = month, 
    values_from = calls
  )
```

