---
title: "Lecture 03: R Functions and Packages"
subtitle: "Getting Started"
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2024-10-09"
date-modified: "2024-10-11"
sidebar: false
execute: 
  echo: true
  warning: false
  eval: true
output-location: default
code-annotations: below
highlight-style: "nord"
format: 
  uark-revealjs:
    scrollable: true
    chalkboard: true
    embed-resources: false
    code-fold: false
    number-sections: false
    footer: "ESRM 64503 - Lecture 03: Object/Function/Package"
    slide-number: c/t
    tbl-colwidths: auto
    output-file: slides-index.html
---

# R Function

## Prebuilt functions

-   **Functions**: Once we defined the objects, the data analysis process can usually be described as [a series of *functions*]{.underline} applied to the data.

    -   In other words, we considered "function" as a set of pre-specified operations (e.g., macro in SAS)

    -   R includes several **predefined functions** and most of the analysis pipelines we construct make extensive use of these.

    -   We already used or discussed the `install.packages`, `library`, and `ls` functions. We also used the function `sqrt` to solve the quadratic equation above.

-   **Evaluation**: In general, we need to use parentheses followed by a function name to evaluate a function.

    -   If you type `ls`, the function is not evaluated and instead R shows you the code that defines the function. If you type [`ls()`](https://rdrr.io/r/base/ls.html) the function is evaluated and, as seen above, we see objects in the workspace.

-   **Function Arguments:** Unlike `ls`, most functions require one or more *arguments* to specify the settings of the function.

    -   For example, we assign different object to the argument of the function `log`. Remember that we earlier defined `coef_a` to be 1:

```{r}
#| eval: true
log(8)
```

## Play sounds

```{r}
#| eval: false
system("say ESRM6 9 9 0 Vi is exciting ")
system("say Professor Jihong Zhang can play the R code all day")
for (i in 1:10) {
  system("say hate you")
}
```

## Want to watch firework in R?

```{r}
#| eval: false
# install.packages('robservable')
library(robservable)
robservable("@rstata/happynewyear", hide = c(1:2, 4:20))
```
