---
title: "Exercise 03: R Functions"
subtitle: ""
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2025-08-18"
sidebar: false
execute: 
  warning: false
  message: false
  eval: false
  echo: true
format: 
  html: 
    page-layout: full
toc: true
toc-depth: 2
lightbox: true
code-fold: show
---

## Creating Custom Functions

1.  Write a function called `fahrenheit_to_celsius` that converts a temperature from Fahrenheit to Celsius.

-   Formula: $C = \frac{(F - 32) \times 5}{9}$
-   Test your function with $F = 98.6$.

```{r}
fahrenheit_to_celsius <- function(temp_f) {
  return((temp_f - 32) * 5 / 9)
}
fahrenheit_to_celsius(temp_f = 98.6)
```

2.  Create a function called `circle_area` that calculates the area of a circle given its radius.

-   Formula: $A = \pi \times r^2$
-   Use the built-in constant `pi` in R and test the function with $r = 5$.

```{r}
# r is the circle radius
circle_area <- function(r){
  return(pi * r^2) 
}
circle_area(r = 5)
circle_area(5) # 5 assign to r
```

------------------------------------------------------------------------

## Piping

1.  Transform the following code into a pipe-based version:

```{r}
result <- filter(mtcars, hp > 100)
result <- arrange(result, desc(mpg))
result <- select(result, hp, mpg)
result
```

```{r}
mtcars |> 
  filter(hp > 100) |> 
  arrange(desc(mpg)) |> 
  select(hp, mpg)
```

------------------------------------------------------------------------

## Debugging Functions

1.  Identify and correct the mistake in the following code:

```{r}
df |> mutate(
  a = (a - mean(a, na.rm = TRUE)) / sd(a),
  b = (b - mean(a, na.rm = TRUE)) / sd(b) # a should be changed to b
)
```

2.  Rewrite the code by creating a custom function `standardized()` to avoid code repetition.

```{r}
standardized <- function(x){
  return((x - mean(x, na.rm = TRUE)) / sd(x))
}
df |> mutate(
  a = standardized(a),
  b = standardized(b)
)
```

------------------------------------------------------------------------

## Arguments in Functions

1.  Create a function `greet` that takes two arguments, `name` (default = "World") and `punctuation` (default = "!"). The function should return a greeting message.

-   Example: `greet("Alice", "?")` should return `"Hello, Alice?"`.

```{r}
greet <- function(name = "World", punctuation = "!") {
  paste0("Hello, ", name, punctuation)
}
greet(name = "Alice", punctuation = "?")
```

2.  Test the `sum()` function by passing named arguments (`x = 3, y = 5`) and explain how the function handles them.

```{r}
sum(x = 3, y = 5)
sum(3, 5)
sum(c(3, 5))
# mean(x = 3, 5)
# mean(c(3, 5))
```

------------------------------------------------------------------------

## Function Scope

1.  Explain why the following code throws an error:

```{r}
add <- function(x, y) {
  result <- x + y # beacuase result is in the local scope
  return(result)
}
add(5, 3)
result  # Error
```

2.  Modify the code to make the variable `result` accessible outside the function.

```{r}
add <- function(x, y) {
  result <<- x + y # beacuase result is in the local scope
  return(result)
}
add(6, 3)
result
```

------------------------------------------------------------------------

## Flexible Arguments

1.  Write a function `multiply_all()` that accepts a variable number of arguments and multiplies all the numbers together. Test your function with `multiply_all(2, 3, 4)`.

```{r}
multiply_all <- function(x, y, z) { # flexible arguments
  x * y * z
}
multiply_all(2, 3, 4)

```

```{r}
multiply_all_flexible <- function(...) {
  prod(c(...))
}
multiply_all_flexible(1, 2, 3)
multiply_all_flexible(1, 2, 3, 5)
multiply_all_flexible(1, 2, 3, 5, 8)
multiply_all_flexible(c(1, 2, 3, 5, 8))
```

1.  Investigate the `mean()` function by using the help page. Explain why it can accept additional arguments like `y` in the code below:

```{r}
mean(x = c(1, 2, 3), y = 3) # because y is flexible argument. It is not included into calculation of mean
```

------------------------------------------------------------------------

## Nested Functions

1.  Write a function `rect_prism_volume()` that calculates the volume of a rectangular prism given its length, width, and height. Inside this function, create another function `rect_area()` to calculate the base area.

    -   The **formula for the volume of a rectangular prism** is:

        $V =  A_b \times h$

        $A_b = l \times w$

        where:

        -   V = Volume of the rectangular prism,

        -   $A_b$ = base area

            -   l = Length,

            -   w = Width,

        -   h = Height.

```{r}
# the first function
rect_area <- function(l, w){
  A_base = l * w
  return(A_base)
}

# the second function that contains the first function
rect_prism_volume <- function(l, w, h){
  return(rect_area(l, w) * h)
}
```

1.  Test your function with length = 5, width = 3, height = 4.

```{r}
rect_prism_volume(l =5, w = 3, h = 4)
```

------------------------------------------------------------------------

## Rescale Function

1.  Create a custom function `rescale()` that rescales a numeric vector to the range 0 to 1.

-   Formula: $x_{\text{scaled}} = \frac{x - \min(x)}{\max(x) - \min(x)}$

```{r}
rescale <- function(x){
  res = (x - min(x)) / (max(x) - min(x))
  return(res)
}
```

2.  Use your `rescale()` function to rescale the columns of the following data frame:

```{r}
set.seed(1234)
df <- tibble(
 a = rnorm(5),
 b = rnorm(5),
 c = rnorm(5)
)
df
```

```{r}
df |> 
  mutate(
    a = rescale(a),
    b = rescale(b),
    c = rescale(c)
  )
```
