---
title: "Example 02: Hypothesis Testing"
subtitle: "Experimental design in Education"
author: "Jihong Zhang*, Ph.D"
institute: | 
  Educational Statistics and Research Methods (ESRM) Program*
  
  University of Arkansas
date: "2025-08-18"
sidebar: false
format: 
  #html: 
  #  page-layout: full
  #  toc: true
  #  toc-depth: 2
  #  lightbox: true
  uark-revealjs:
    chalkboard: true
    embed-resources: false
    code-fold: true
    number-sections: true
    number-depth: 1
    footer: "ESRM 64503: Lecture01"
    slide-number: c/t
    tbl-colwidths: auto
    output-file: slides-index.html
    scrollable: true
#jupyter: python3
---

## Research Question & Data

**Context:**

-   Data were collected for two species: *O. exclamationis* and *O. niveus*. The data are contained in a data frame called `crickets` with a total of 31 data points.

-   Study on cricket chirp rates and temperature for two species

-   Data: Temperature (°C) and chirps per minute

**Research Questions:**

-   Does temperature affect chirp rate?

-   Do species differ in their chirp rates?

-   Do species respond differently to temperature?

```{r setup}
library(tidyverse) #install.packages("tidyverse")
data(crickets, package = "modeldata")
```

## Data Visualization

```{r cricket_plot}
ggplot(crickets, 
       aes(x = temp, y = rate, color = species)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Temperature (C)", 
       y = "Chirp Rate (per minute)")
```

## Statistical Hypotheses & Model

**Null Hypotheses:**

1.  Temperature has no effect on the chirp rate.
2.  There are no differences between the species’ chirp rate, e.g., overall chirp rate and sensitivity to temperature

**Null Hypotheses for hypothesis testing:**

1\. H₀: Temperature has no effect ($β_{temp}$ = 0)

2\. H₀: No species difference ($β_{species}$ = 0)

3\. H₀: No interaction ($β_{temp×species}$ = 0)

```{r model_fit}
# Full model with interaction
interaction_fit <- lm(rate ~ (temp + species)^2, data = crickets)
```

## Model Diagnostics

```{r diagnostics}
par(mfrow = c(1, 2))
plot(interaction_fit, which = 1) # Residuals vs Fitted
plot(interaction_fit, which = 2) # Normal Q-Q
```

## Testing Interaction Effect

```{r test_interaction}
# Fit main effects model
main_effect_fit <- lm(rate ~ temp + species, data = crickets)

# Compare models
anova(main_effect_fit, interaction_fit)
```

**Result:** Interaction not significant (p \> 0.05). Choose simpler model without interaction

## Final Model Results

```{r final_model}
summary(main_effect_fit)
```

**Interpretation:**

-   Temperature effect: +3.75 chirps/°C (p \< 0.001)

-   O. niveus: 17 fewer chirps/minute (p \< 0.001)

## Model Predictions

```{r predictions}
new_temps <- data.frame(
  species = "O. exclamationis",
  temp = seq(15, 25, by = 5)
)

predict(main_effect_fit, new_temps)
```

## Conclusions

**Statistical Evidence:**

1\. Strong temperature effect (p \< 0.001)

2\. Clear species difference (p \< 0.001)

3\. Similar temperature response (no interaction)

**Limitations:**

-   Valid within observed temperature range (15-32°C)

-   Only two species studied

## Takeaway Note: the key steps

Here are the key steps for performing inferential statistics using the cricket example:

1.  Data visualization and propose Null Hypotheses
    -   Temperature has no effect on chirp rate
    -   No differences between species' chirp rates
2.  Fit Initial Model
    -   Used `lm()` with interaction terms between temperature and species
    -   Formula: `rate ~ (temp + species)^2`
3.  Check Model Diagnostics
    -   Examined residual plots vs predicted values
    -   Checked normality of residuals using Q-Q plot
    -   Verified assumptions were reasonable
4.  Test Model Components
    -   Compared models with/without interaction using ANOVA
    -   Found interaction term was not statistically significant (p \> 0.05)
    -   Selected simpler main effects model
5.  Interpret Final Model
    -   Analyzed **coefficients** and **p-values** for temperature and species effects
    -   Temperature showed significant positive effect
    -   Species showed significant difference in baseline chirp rates
    -   Limited conclusions to observed temperature range
