---
title: "Error Handling in R"
subtitle: "How to handling error of simulation study and functions in R"
description: "Error handling is important for simulation study"
author: 'Jihong Zhang'
date: 'August 31 2024'
sidebar: false
categories:
  - R
  - Simulation
execute: 
  eval: true
  echo: true
  warning: false
  error: true
format: 
  html: 
    code-fold: false
    code-summary: 'Click to see the code'
    number-sections: true
bibliography: references.bib
---

> This blog is based on @andersonErrorHandlingGeneration2020 and @wickhamConditionsAdvanced2024.

## What is Error in R?

In R, errors are signaled, or thrown, by `stop()` or `stopifnot()`:

```{r}
stop("Something erroneous has occurred!")
stopifnot(5 < 0)
```

`stopifnot()` is basically a wrap-up function of `if`-condition and `stop` to detect if any of the expressions are not all TRUE.

Sometimes, error messages are meaningful and deliberately created/controlled by function authors while it is meaningless and subjected to be controlled by R users. Here are some scenairos:

**Example 1** **- Uncontrolled error**: You want to ran a `for` iteration while there is some error message in some iterations. The error message contains nothing but the stopping signals from data analysis for certain iterations. This type of error message make not much sense to understanding why this error occur.

**Example 2 - Controlled error**: You are using a function inappropriately and this function throws you some message. If the function is well coded, the error message does tell you the "**Problem statement**" using `must` or `can't`:

```{r}
#| error: true
#| results: hold
dplyr::nth(1:10, 1:2)
as_vector(environment())
```

Some functions also provide hints when error occurs:

```{r}
#| error: true
#| results: hold
dplyr::filter(iris, Species = "setosa")
```

## Functions for error handling in R

> Condition handling tools, like `withCallingHandlers()`, `tryCatch()`, and `try()` allow you to take specific actions when a condition occurs. For example, if you’re fitting many models, you might want to continue fitting the others even if one fails to converge.

### Use `try` to ignore errors

With the `try` function, you can "hide" or ignore error signals to continue the execution:

```{r}
## Function without try
fun0 <- function() {
  x <- b
  x <- 3
  print(x)
}
fun0()
## Function with try
fun1 <- function() {
  try(x <- b)
  x <- 3
  print(x)
}
fun1()

## Multiple errors in one function
options(error = function() traceback(3))
fun1 <- function() {
  try(x <- b1) # This will throw error but will keep the execution
  try(x <- 3) # This will not throw error and will execute the code normally
  print(x)
  x <- b2 # This will also throw error but will stop the execution
  print(x)
}
fun1()
```

The function with `try` does not stop executing the rest of the syntax after error occurs (`x <- b`). However, there is one issue of `try()`: it only ignores single error message at once and you have to know where the error comes from beforehand.

In practice, you can ignore error when running the simulation:

```{r}
for (x in list(1, 2, "not a number")) {
  y <- 3*x
  print(paste0("3x= ", y))
}

for (x in list(1, 2, "not a number")) {
  # when x iterate over thirt element, try() ignore the error
  # and still keep y = 2
  try(y  <- 3*x, silent = TRUE) 
  try(print(paste0("3x= ", y)))
}
```

If you want to assign b to x if b is available and assign NA to x if b is unavailable, you can:

1.  Use `if` condition to detect whether b is available or not;
2.  Use `tryCatch()` to use condition name (error, warning) as parameter

### Use `tryCatch` to handle errors

`tryCatch()` is a general tool for handling conditions: in addition to errors, you can take different actions for warnings, messages, and interrupts [@wickhamConditionsAdvanced2024].

```{r}
# when b is unavailable
tryCatch({
  x <- b
}, error = function(e) {x <<- NA})
x

# when b is available
b <- 3
tryCatch({
  x <- b
}, error = function(e) {x <<- NA})
x
```

The `tryCatch()` will catch the error condition (warning) and execute another function you defined (`error = function(e)` in the example). Note that if you want to change x's value in `error` function when `x<-b` has error, you need to use `<<-` assign symbol to transfer x value into [global environment](https://adv-r.hadley.nz/environments.html).

### Use `withCallingHandlers` to handle errors

```{r}
withCallingHandlers({
  x <- b2
}, error = function(e) {x <<- NA})
print(x)
```
