---
title: Simulation Study of Linking Using Mirt
author: Jihong Zhang
date: '2017-11-10'
categories:
  - R
  - mirt
  - linking
format: 
  html:
    toc: true
    code-fold: true
    code-summary: ' R Code'
    code-line-numbers: true
    number-sections: true
    number-offset: 1
---


> This simulation study is to show how to do IRT Linking Process using mirt R Package. The simulation data includes 2 forms - Form A and Form B. These 2 forms are simulated based on 2 groups of individual, one group has 0 mean trait, another has 0.25 mean trait. Both groups have same sd. <!--more-->

# Introduction

The means and sds of simulated Form A and B are like:

$$
\theta_{A} = \theta_{B} - 0.25 \\
\sigma_A^2 = \sigma_B^2
$$

# Calibration of Form A

The mean of $\theta$ for individuals administrated with form A is 0, the standard deviation (SD=1). In the dataset, X is ID, V1 is true trait ($\theta$), V3 to V52 is unique items, V54 to V63 are common items.

## Look at the data

First of all, have a look at the data

```{r , message=FALSE}
library(mirt)
library(tidyverse)
library(knitr)
### Read in Raw data from Form A:
dat <- read.csv(file="FormA.csv")
glimpse(dat)
```

## Plot the density of true $\theta$ of Group A

From the density function, $mu_{\theta}$ is 0, $sd_{\theta}$ is 1.

```{r theta}
plot(density(dat$V1), main = "Group A True Trait Density", 
     xlab=expression(theta) )
```

## CTT Table

CTT table could provide a brief description of table. 2 key valables in the table is item difficulty (*item.diff*) calculated by item means $P(y=1)$ and item discrimination (item.disc), which is item-total correlation.

### Clean data

By cleaning, data has 60 items including 50 unique item (from item1 to item50) and 10 common items (from item51 to item60). The sample size is 5000.

```{r}
dat_cali <- dat %>% select(V3:V52, V54:V63)
colnames(dat_cali) <- paste0("item",1:60)
N <- nrow(dat_cali)
n <- ncol(dat_cali)
```

### Classical Test Theory

Then calculate the CTT table for Form A. The item discrimination and difficulty could be compared between Form A and Form B. Because the relationship between trait and total score is non-linear, so there is effect of shrinkage.

```{r}
# item stats
## item discrimnation
item.disc <- apply(dat_cali, 2, function(x) cor(x, rowSums(dat_cali, na.rm = TRUE)))

## item difficulty
item.diff <- colMeans(dat_cali)

## item response frequency
item.freq <- reduce(lapply(dat_cali, table), bind_rows)

CTT <- cbind(item.disc, item.diff, item.freq)

kable(CTT, digits = 3, caption = "CTT Table for Form A")
```

## Final Calibration of Form A

### Model Specification

```{r , message=FALSE, results='hide'}
SPECS <- mirt.model('F = 1-60
                    PRIOR = (1-60, a1, lnorm, 0, 1),
                    (1-60,  d,  norm, 0, 1),
                    (1-60,  g,  norm, -1.39,1)')
mod_A3PL <- mirt(data=dat_cali, model=SPECS, itemtype='3PL')
parms_a <- coef(mod_A3PL, simplify=TRUE, IRTpars = TRUE)$items
```

```{r , message=FALSE}
a_A <- parms_a[,1] 
b_A <- parms_a[,2] 
c_A <- parms_a[,3] 
theta_A <- fscores(mod_A3PL,method="EAP",
                   full.scores=TRUE, full.scores.SE=TRUE,
                   scores.only=TRUE)


```

Using 3-PL for irt model of form A. Extracting the cofficients (a, b, c) of IRT. The model-implied theta was outputed

The plot below suggest that SE is low when theta is close to mean, but low theta and high theta has large SE.

```{r SE}
plot(theta_A[,1],theta_A[,2])
```

# Calibration of Form B

The structure of Form B is as same as Form A.

```{r}
dat_b <- read.csv(file="FormB.csv")
glimpse(dat_b)
```

```{r theta_b}
library(ggplot2)

ggplot(data = dat_b) +
  ggtitle("Group B True Trait Density") +
  geom_density(aes(x = V1), fill = "skyblue", col = "skyblue", alpha=0.8) +
  xlab("True Latent Trait")
  
```

```{r stats2}
dat_cali2 <- dat_b %>% select(V3:V52, V54:V63)
colnames(dat_cali2) <- paste0("item",1:60)
N <- nrow(dat_cali2)
n <- ncol(dat_cali2)


# item stats
## item discrimnation
item.disc <- apply(dat_cali2, 2, function(x) cor(x, rowSums(dat_cali, na.rm = TRUE)))

## item difficulty
item.diff <- colMeans(dat_cali2)

## item response frequency
item.freq <- reduce(lapply(dat_cali2, table), bind_rows)

CTT2 <- cbind(item.disc, item.diff, item.freq)

kable(CTT2, digits = 3, caption = "CTT Table for Form A")

```

## Final Calibration of Form A

### Model Specification of B

```{r , message=FALSE, results='hide'}
SPECS2 <- mirt.model('F = 1-60
                    PRIOR = (1-60, a1, lnorm, 0, 1),
                    (1-60,  d,  norm, 0, 1),
                    (1-60,  g,  norm, -1.39,1)')
mod_B3PL <- mirt(data=dat_cali2, model=SPECS, itemtype='3PL')
parms_b <- coef(mod_B3PL, simplify=TRUE, IRTpars = TRUE)$items
```

```{r , message=FALSE}
a_B <- parms_b[,1] 
b_B <- parms_b[,2] 
c_B <- parms_b[,3] 
theta_B <- fscores(mod_B3PL,method="EAP",
                   full.scores=TRUE, full.scores.SE=TRUE,
                   scores.only=TRUE)
head(theta_B, 20) %>% kable(digits = 3, caption = "Model-implied Theta of B")
```

## b-plot

The relationship between b parameters of A and B reflect the latent traits of A and B:

$$
\theta_{A} = \theta_{B} - 0.25 \\
\sigma_A^2 = \sigma_B^2
$$ thus, $b_B -b_A$ should also be -0.25. the estimated difference of b is calculate by the mean of b parametes of Form A's common items and that of Form B's common items. Thus, it is very close to difference of true traits.

```{r}
### b-plot
plot(b_A[51:60],b_B[51:60],
     main=paste0("r =", round(cor(b_A[51:60],b_B[51:60]),5)), 
     xlab = "b_A",
     ylab = "b_B"
     )
# mean(b_B[51:60])-mean(b_A[51:60])
```

## a-plot

Because $\theta_B - \theta_A = 0.25$, so a parametes are: $$
a_A / a_B = 1
$$

The true ratio of (means of) a parameters is 1.016284, which is very close to 1. SD of A and B are both close to 1.

```{r}
### a-plot
plot(a_A[51:60],a_B[51:60],
     main=round(cor(a_A[51:60],a_B[51:60]),5)
     )

# mean(a_A[51:60]) / mean(a_B[51:60]) 


#SDs of b-values across forms
SD_bA <- sd(b_A[51:60])
SD_bB <- sd(b_B[51:60])

Mean_bA <- mean(b_A[51:60])
Mean_bB <- mean(b_B[51:60])
```

# Linking

```{r}

###
### Run one or the other, NOT BOTH!
###

### MS linking: place item parameters from
### Form B on the scale of Form A
slope <- SD_bA / SD_bB
inter <- Mean_bA - slope*Mean_bB

### MM linking: place item parameters from
### Form B on the scale of Form A
# slope <- Mean_aA / Mean_aB
# inter <- Mean_bA - slope*Mean_bB


###
### Perform the Linking
###

LINKED_items <- matrix(0,50,3)

#Column 3 is c, it stays the same
LINKED_items[,3] <- c_B[1:50]

#Column 2 is b, it is linked:
LINKED_items[,2] <- b_B[1:50]*slope + inter

#Column 1 is a, it is also linked:
LINKED_items[,1] <- a_B[1:50] / slope

### Now the ITEM BANK has 110 items all
### linked to a common metric!
rownames(LINKED_items) <- paste0("item_B", 1:50)
ITEM.BANK <- rbind(parms_a[,-4], LINKED_items)
# write.csv(ITEM.BANK,file="item_bank.csv")

#Link the theta estimates
LINKED_theta <- theta_B[,1]*slope + inter
LINKED_se <- theta_B[,2]/slope

# write.csv(cbind(LINKED_theta,LINKED_se),file="LINKED_FormB_theta_est.csv")
paste0("Mean of Theta of B is ",mean(theta_B[,1]) %>% round(3))
paste0("SD of Theta of B is ", sd(theta_B[,1]) %>% round(3))


# after Linked
print("After Linking:")

paste0("Mean of Theta of B is ",mean(LINKED_theta) %>% round(3))
paste0("SD of Theta of B is ",sd(LINKED_theta) %>% round(3))

# Theta of A is
paste0("Mean of Theta of A is ",mean(theta_A[,1]) %>% round(3))
paste0("SD of Theta of A is ", sd(theta_A[,1]) %>% round(3))

```
