---
title: Which Country own the most Liberty Nobel Prizes? France? Ireland?
authors: "Jihong Zhang"
date: '2019-05-20'
categories:
  - visualization
  - blog
  - ggplot2
---

> It is of my interest to use `ggplot` package to visualize some fun data. In this post, I tried to play with a Nobel Prizes Data including countries, prize year, each prize. The goal is to plot a cumulative traceplot of Liberty Nobel Prizes for top 10 countries.

# Load Packages

`tidyverse` package include some very useful tools such as `ggplot2`, `tidyr` and `dplyr`.

```{r message=FALSE, warning=FALSE,  paged.print=FALSE}
library(tidyverse)
library(LaCroixColoR)
library(ggthemes)
library(ggimage)
```

`ggimage` package was used to add country flags to ggplot layer. `LaCroixColoR` package used for selecting different colors.

# Data

Import the data.

```{r}
nobel_winners <- read_csv("data_2019-05-14.csv",
                          col_types = "dccccdccDccccccDcc")
str(nobel_winners)
unique(nobel_winners$category)
```

`tribble` function allows to create a table by columns. `~country` will generate a new columns.

```{r}
df_countrycode <-
  tribble(~country, ~code, 
          "United States of America", "US",
          "Germany", "DE",
          "United Kingdom", "GB",
          "France", "FR",
          "Japan", "JP",
          "Netherlands", "NL",
          "Sweden", "SE",
          "Russia", "RU",
          "Canada", "CA",
          "Austria", "AT",
          "Spain", "ES",
          "Denmark", "DK",
          "Ireland", "IE",
          "Italy", "IT"
          )
```

# Tables

`pull()` function is similar to use `[]` to return the value of one column.

```{r}
countries <- 
  nobel_winners %>% 
  filter(category == "Literature") %>% 
  count(birth_country, sort = T) %>% 
  head(n = 10) %>% # Top 10 countries
  pull(birth_country)
```

Adding some columns such as the total number of Liberty Prizes (`n_prize`), the first year of prize, the last year and the cumulative prizes.

```{r}
nobel_countries <- 
  nobel_winners %>% 
  filter(birth_country %in% countries, category == "Literature") %>% 
  select(prize_year, birth_country) %>% 
  arrange(prize_year) %>% 
  group_by(birth_country) %>% 
  mutate(n_prize = n(),
         first_prize = min(prize_year),
         last_prize = max(prize_year),
         cum = row_number()
  ) %>% 
  ungroup()
nobel_countries
```

```{r}
nobel_countries <- 
  nobel_countries %>% 
  filter(cum == 1) %>% 
  mutate(cum = 0) %>% 
  bind_rows(nobel_countries) %>% 
  arrange(prize_year, cum) %>% 
  mutate(birth_country = fct_reorder(birth_country, n_prize, .desc = TRUE))
nobel_countries
```

```{r}
first_last_nobel <- 
  nobel_countries %>% 
  select(birth_country, n_prize, first_prize, last_prize) %>% 
  mutate(birth_country = as.character(birth_country)) %>% 
  distinct() %>% 
  left_join(df_countrycode, by=c("birth_country" = "country"))
first_last_nobel
```

# Plot

```{r}
pp1 <- ggplot(nobel_countries) +
  aes(x = prize_year, y = cum, group = birth_country) +
  geom_line(aes(color = birth_country)) +
  geom_point(data = first_last_nobel, y = 0,
             aes(x = first_prize, color =birth_country)) +
  geom_flag(data = first_last_nobel, size = 0.03, asp= 2,
            aes(x= last_prize, y = n_prize, image = code)) +
  scale_color_manual(values = lacroix_palette("PassionFruit", n = 10, type = "continuous")) +
  scale_y_continuous(limits = c(NA, 12), breaks = c(0,3,6,9,12)) +
  labs(title = "Number of Liberty Nobel Prizes by Countries",
       color = NULL,
       caption = "Source: The Nobel Prize\n@_abichat for #TidyTuesday") +
  theme_wsj(color = "gray") +
  theme(legend.position = "bottom",
        plot.caption = element_text(size = 10, family = "Georgia"),
        plot.title = element_text(size = 18, family = "Andale Mono"),
        legend.text =  element_text(family = "Georgia")
        )

pp1
# ggsave("~/图片/plot_2019-05-14.png", width = 29, height = 21, units = "cm", dpi = "retina")

```

# Animation

Finally, let's use `gganimate` package to add some animation.

```{r , cache = TRUE}
library(gganimate)
pp1 +
  transition_reveal(prize_year)
```
